---
title: "Initial data presentation"
author: "callumcoode"
date: "23/03/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## First presentation of my data set

I wanted to find something related to Covid-19 but looking at something positive. 
I found data from [WHO](https://apps.who.int/flumart/Default?ReportNo=12) on cases of influenza as I have heard from many places that people haven't got flu or colds as they usually would due to social distancing.

I have processed the data to just look at the total cases of all types of flu on a weekly basis. From initially looking at the data, it does look like the cases in 2020/2021 are a lot lower.

```{r library set up, include=FALSE}
library(here)
library(tidyverse)
```

```{r processing data}
df <- read.csv(here("data", "rawfludata.csv"))

df <- df %>% 
  rename(year = Year, week = Week, start = Start.date, end = End.date, cases = Total.number.of.influenza.positive.viruses)

df <- df %>% 
  select(year, week, start, end, cases)

head(df)
```
