---
title: "First presentation of my data set"
author: "Callum Coode"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Effect of social distancing on cases of flu

I wanted to find something related to Covid-19 but looking at something positive. 
I found data from *[WHO on cases of influenza](https://apps.who.int/flumart/Default?ReportNo=12)* as a lot of people are saying that they haven't got flu or colds this year due to social distancing.
I have processed the data to just look at the total cases of all types of flu on a weekly basis.


```{r library set up, include=FALSE}
library(here)
library(tidyverse)
```

```{r processing data}
df <- read.csv(here("data", "rawfludata.csv"))

df <- df %>% 
  rename(year = Year, week = Week, start = Start.date, end = End.date, 
         cases = Total.number.of.influenza.positive.viruses)

df <- df %>% 
  select(year, week, start, end, cases)

head(df, 10)
```


From initially looking at the data, it does look like the cases in 2020/2021 are a lot lower. For example, the mean number of cases in 2020 is `r round(mean(df[df$year == 2020, "cases"]), digits = 2)`, whereas in 2019 the mean is `r round(mean(df[df$year == 2019, "cases"]), digits = 2)`.


```{r line graph}
df$year <- as.character(df$year)

p1 <- ggplot(df, aes(x = week, y = cases, group = year, colour = year))
p1 + geom_line() + geom_point()
```

Since cases are usually a lot higher in the winter and lockdown started on 23rd March (week 13), I may display this data from week 13 to week 13 to better show the impact of social distancing on cases. I may also include bar graphs showing highest or total cases for each of these periods.

**Please note this does not include data from 2021.**
There is currently no data for the most recent 2 weeks of 2021, but this will be added as soon as this becomes available, hopefully in the next few weeks.

